---
title: "GBM_01"
author: "Teresa G Krieger"
date: "2/20/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### GBM analysis 01: clustering of all cells

##### Load libraries and initialise
```{r echo=FALSE}

# ---- Load packages ----
library(Seurat)
library("RColorBrewer") 
library(data.table)
library(plyr)
# 

# --- Set working directory ----
folderpath <- "~/Documents/Postdoc/GBM/Analysis_scRNA-seq/" 
setwd(folderpath)

# ---- User-defined functions ----
printf <- function(...) cat(sprintf(...))

# ---- Initialise variables ----
sids <- c("NSC", "F2m", "F3m", "F6m", "F9m", "F2", "F3", "F6", "F9")
PatientNames <- c("F2", "F3", "F6", "F9")
SampleNames <- c("NSC", PatientNames)
ChipIDs <- c("99131", "99205", "99209", "99165", "99132", "97659", "97668", "97990", "97911")
colors <- brewer.pal(10, "Set3")
cds <- read.csv(paste0(folderpath,"resources/GRCh37_protein_coding.csv"))[,1]

```

##### Load count data, normalise, filter based on mito.genes
```{r echo=FALSE}

# ---- Load count data ----

# Read in raw counts
count.matrix <- fread(file = paste0(folderpath, "rawdata/AllSamples_counts_noMeanCen_noLogTrans.csv"),
                      header = T, 
                      data.table = T
                      )    
c <- count.matrix
all.gene.names <- c[,1]
c[,1] <- NULL 
count.matrix <- as.matrix(c)
rownames(count.matrix) <- t(all.gene.names)
rm(c)

# Create Seurat object from full count.matrix
gbm <- CreateSeuratObject(count.matrix, 
                          project = "gbm", 
                          min.cells = 0, min.genes = 0, is.expr = 0, 
                          normalization.method = NULL,
                          do.scale = FALSE, do.center = FALSE,
                          names.field = 1, names.delim = "-"
)

# Calculate fraction of mitochondrial genes
mito.genes <- grep(pattern = "^MT-", x = rownames(x = gbm@raw.data), value = TRUE)
percent.mito <- Matrix::colSums(gbm@raw.data[mito.genes,])/
                Matrix::colSums(gbm@raw.data)
gbm <- AddMetaData(object = gbm, metadata = percent.mito, col.name = "percent.mito")
```

##### Fraction of mitochondrial genes
```{r echo=FALSE}
VlnPlot(object = gbm, features = "percent.mito", point.size.use = 0.1, y.max = 0.2)
VlnPlot(object = gbm, features = "nUMI", point.size.use = 0.1)
VlnPlot(object = gbm, features = "nGene", point.size.use = 0.1, y.max = 200)
```

##### Filter cells with <150 genes or > 15% mito.genes
```{r echo=FALSE}
# ---- Filter cells with <150 genes or > 15% mito.genes ----
gbm <- FilterCells(object = gbm, subset.names = c("nGene", "percent.mito"), 
                   low.thresholds = c(150, -Inf), high.thresholds = c(Inf, 0.15))
# retain only filtered cells in count.matrix
count.matrix <- count.matrix[,colnames(gbm@data)]

# ---- LogNormalise data ---- log1p(value/colSums[cell-idx]*scale_factor)
gbm <- NormalizeData(object = gbm, 
                     normalization.method = "LogNormalize", 
                     scale.factor = 10000)
```

##### Exclude genes with low aggregate expression 
```{r eval=FALSE, echo=FALSE}
# Calculate aggregate expression of each gene
scale <- colSums(gbm@raw.data)/1e6
cpm <- matrix(ncol = ncol(gbm@raw.data), nrow = nrow(gbm@raw.data))
for (i in 1:length(scale)){
  cpm[,i] <- gbm@raw.data[,i]/scale[i]
}
colnames(cpm) <- colnames(gbm@raw.data)
rownames(cpm) <- rownames(gbm@raw.data)
AggregateExpression <- log(rowMeans(cpm) + 1)
names(AggregateExpression) <- rownames(gbm@raw.data)

# Exclude genes with aggregate expression below 2 --> 8533 genes 
genes.use <- (AggregateExpression>2)
full <- gbm@raw.data
gbm@raw.data <- gbm@raw.data[genes.use,]
gbm@data <- gbm@data[genes.use,]
sum(genes.use)
```

##### Dimensional reduction and cluster identification
```{r echo=FALSE}
# ---- Dimensional reduction ----

# Calculate relative expression for each gene in each sample, compared to all samples
# with regression against nUMI and percent.mito
gbm <- ScaleData(gbm,
                 vars.to.regress = c("nUMI", "percent.mito"),
                 do.scale = T, model.use = "linear", use.umi = F, scale.max = 10,
                 do.center = T,
                 block.size = 1000,
                 min.cells.to.block = 10000,
                 display.progress = TRUE
)

# Run PCA on all genes
gbm@var.genes <- rownames(gbm@data)
gbm <- RunPCA(object = gbm, pc.genes = gbm@var.genes, do.print = TRUE, pcs.print = 1:10, 
              genes.print = 20)
gbm <- ProjectPCA(object = gbm, do.print = TRUE, pcs.print = 1:10, 
                  genes.print = 20)
# PCA plots
VizPCA(object = gbm, pcs.use = 1:2)
PCAPlot(object = gbm, dim.1 = 1, dim.2 = 2, group.by="orig.ident")

# Select sigPC
PCHeatmap(object = gbm, pc.use = 1:6, cells.use = 500, do.balanced = F, 
          label.columns = FALSE, use.full = FALSE)
PCHeatmap(object = gbm, pc.use = 7:12, cells.use = 500, do.balanced = F, 
          label.columns = FALSE, use.full = FALSE)
PCHeatmap(object = gbm, pc.use = 13:18, cells.use = 500, do.balanced = F, 
          label.columns = FALSE, use.full = FALSE)
PCElbowPlot(object = gbm)
sigPC = 9

# Run tSNE
set.seed(123)
gbm <- RunTSNE(object = gbm, dims.use = 1:sigPC, do.fast = TRUE)

# tSNE plot
TSNEPlot(object = gbm, group.by = "orig.ident")

# Identify clusters (resolution 0.4)
gbm <- FindClusters(object = gbm, reduction.type = "pca", dims.use = 1:sigPC, 
                    resolution = 0.4, print.output = F, save.SNN = T, force.recalc = T)
ClusterID <- gbm@meta.data$res.0.4
# manually divide the F2 cluster into 'mixed' and 'unmixed'
ClusterID[
  as.logical((gbm@meta.data$ClusterID=="2")*
               (gbm@meta.data$orig.ident=="F2"))] <- 9
gbm@meta.data$ClusterID <- ClusterID

# ---- Save results of dimensional reduction ----

# save gene loadings of all genes in PCs 1-10
for (pc in 1:10){
  gene.loadings <- gbm@dr$pca@gene.loadings.full[,pc]
  gene.loadings.rank <- gene.loadings
  gene.loadings <- gene.loadings[order(-gene.loadings.rank)]
  if (pc==1){
    gene.loadings.all <- names(gene.loadings)
  } else {
    gene.loadings.all <- cbind(gene.loadings.all, names(gene.loadings))
  }
  write.csv(gene.loadings, file=paste0(folderpath,sprintf("seurat-out/GeneLoadings_full_PC_%s.csv",pc)))
}
write.csv(gene.loadings.all, file=paste0(folderpath,"seurat-out/GeneLoadings_full_allPCs.csv"))

# save gene loadings of selected genes in PCs 1-10
for (pc in 1:10){
  gene.loadings <- gbm@dr$pca@gene.loadings[,pc]
  gene.loadings.rank <- gene.loadings
  gene.loadings <- gene.loadings[order(-gene.loadings.rank)]
  if (pc==1){
    gene.loadings.all <- names(gene.loadings)
  } else {
    gene.loadings.all <- cbind(gene.loadings.all, names(gene.loadings))
  }
  write.csv(gene.loadings, file=paste0(folderpath,sprintf("seurat-out/GeneLoadings_PC_%s.csv",pc)))
}
write.csv(gene.loadings.all, file=paste0(folderpath,"seurat-out/GeneLoadings_allPCs.csv"))

# PCA plot 
plot1 <- PCAPlot(object = gbm, dim.1 = 1, dim.2 = 2, group.by="orig.ident", do.return=T)
plot2 <- PCAPlot(object = gbm, dim.1 = 3, dim.2 = 4, group.by="orig.ident", do.return=T)
plot3 <- PCAPlot(object = gbm, dim.1 = 5, dim.2 = 6, group.by="orig.ident", do.return=T)
filename <- paste0(folderpath, "seurat-out/PCAPlots_12_34_56.pdf")
pdf(filename,width=24,height=7,paper='special') 
plot_grid(plot1, plot2, plot3, labels = c("PC1 vs PC2", "PC3 vs PC4", "PC5 vs PC6"), hjust = -1, vjust = 2, ncol = 3)
dev.off()

# PCA plots colored by orig.ident and ClusterID
plot1 <- PCAPlot(object = gbm, dim.1 = 1, dim.2 = 3, group.by="ClusterID", do.return=T)
plot2 <- PCAPlot(object = gbm, dim.1 = 1, dim.2 = 3, group.by="orig.ident", do.return=T)
filename <- paste0(folderpath, "seurat-out/PCAPlot_byClusterID_bySampleID_res0p4.pdf")
pdf(filename,width=16,height=7,paper='special') 
plot_grid(plot1, plot2, labels = c("Cluster ID", "Sample ID"), hjust = -1, vjust = 2, ncol = 2)
dev.off()

# QC on PCA plot 
filename <- paste0(folderpath, "seurat-out/QC_on_PCAPlot.pdf")
pdf(filename,width=24,height=7,paper='special') 
print(FeaturePlot(gbm, features.plot = c("nGene", "nUMI", "percent.mito"),
                  nCol = 3,
                  min.cutoff = NA, max.cutoff = NA,
                  dim.1 = 1, dim.2 = 3, pt.size = 1,
                  cols.use = c("yellow", "red"), pch.use = 16,
                  reduction.use = "pca",
                  no.legend = F,
                  do.return = T
))
dev.off()

# tSNE plot
plot1 <- TSNEPlot(object = gbm, dim.1 = 1, dim.2 = 2, do.return = TRUE, group.by = "ClusterID", no.legend = F, do.label = F)
plot2 <- TSNEPlot(object = gbm, dim.1 = 1, dim.2 = 2, do.return = TRUE, group.by = "orig.ident", no.legend = F, do.label = F)
filename <- paste0(folderpath, "seurat-out/TSNEPlot_byClusterID_bySampleID_res0p4_afterReg.pdf")
pdf(filename,width=16,height=7,paper='special') 
plot_grid(plot1, plot2, labels = c("Cluster ID", "Sample ID"), hjust = -1, vjust = 2)
dev.off()

# QC on tSNE plot
filename <- paste0(folderpath, "seurat-out/QC_on_tSNE.pdf")
pdf(filename,width=24,height=7,paper='special') 
print(FeaturePlot(gbm, features.plot = c("nGene", "nUMI", "percent.mito"),
                  nCol = 3,
                  min.cutoff = NA, max.cutoff = NA,
                  dim.1 = 1, dim.2 = 2, pt.size = 1,
                  cols.use = c("yellow", "red"), pch.use = 16,
                  reduction.use = "tsne",
                  no.legend = F,
                  do.return = T
))
dev.off()

```

##### Collate meta data and save Seurat object
```{r echo=FALSE}

# ---- Collate meta data and save Seurat object ----

SampleID <- gbm@meta.data$orig.ident
TumorBrain <- mapvalues(ClusterID, 
                        0:9, 
                        c("Brain", "Brain", "Tumor", "Brain", "Tumor", "Tumor", "Tumor", "Tumor", "Tumor", "Tumor"))
MixedUnmixed <- mapvalues(SampleID, 
                          c("F2",  "F2m", "F3",  "F3m", "F6",  "F6m", "F9",  "F9m", "NSC"),
                          c("Unmixed", "Mixed", "Unmixed", "Mixed", "Unmixed", "Mixed", "Unmixed", "Mixed", "Unmixed"))
PatientID <- mapvalues(SampleID, 
                       c("F2",  "F2m", "F3",  "F3m", "F6",  "F6m", "F9",  "F9m", "NSC"),
                       c("F2", "F2", "F3", "F3", "F6", "F6", "F9", "F9", "NSC"))

MetaData <- data.frame(SampleID = SampleID,
                       TumorBrain = TumorBrain,
                       MixedUnmixed = MixedUnmixed,
                       PatientID = PatientID
)

gbm@meta.data <- cbind(gbm@meta.data, MetaData)

# Save Seurat object
gbm.AllCells <- gbm
save(gbm.AllCells, file = paste0(folderpath,"seurat-datasets/SeuratObj_gbm_AllCells_afterTSNE_afterRegression.Rdata"))

```

##### Sanity checks
```{r echo=FALSE}
# ---- Check whether cells in 'tumor' samples are from GBM-only cultures etc.
table(gbm@meta.data$TumorBrain[gbm@meta.data$SampleID %in% c("F2", "F3", "F6", "F9")])
table(gbm@meta.data$SampleID[gbm@meta.data$TumorBrain == "Tumor"])
table(gbm@meta.data$SampleID[gbm@meta.data$TumorBrain == "Brain"])
# --> 95% accurate
```

##### Expression of GBM stem cell markers
```{r echo=FALSE}
# ---- Expression of GBM stem cell markers -----------

pdf(paste0(folderpath, "seurat-out/FeaturePlot_GBMSC_Markers.pdf"), width=10, heigh=10, paper='special')
FeaturePlot(object = gbm, features.plot = c("PROM1", "ITGA6", 
                                            "GFAP", "CD44", 
                                            "EGFR",
                                            "OLIG2", "SOX2",
                                            "NES", "PAX6"),
            cols.use = c("grey", "red"), 
            pt.size = 0.5,
            reduction.use = "tsne")
dev.off()
```


